\chapter*{ABSTRACT}
\addcontentsline{toc}{chapter}{ABSTRACT}

The rapid growth of \textit{Software as a Service} (\textit{SaaS}) applications serving multiple organizations requires an architecture capable of managing multiple tenants efficiently without increasing infrastructure complexity. The \textit{shared database} and \textit{shared schema} approach has emerged as a promising solution by allowing multiple tenants to share the same database and database schema, thereby improving resource utilization. However, this approach presents challenges in maintaining tenant data isolation while supporting automated tenant provisioning. This research aims to design a multitenant architecture based on a \textit{shared database} and \textit{shared schema} for an asset, inventory, and rental management platform integrated with RTPintar. The proposed architecture implements \textit{middleware}-based tenant identification, application-layer tenant scoping, the use of a \textit{tenant\_id} attribute for data access control, and an automated \textit{tenant provisioning} mechanism to support the onboarding of new organizations. The evaluation focuses on infrastructure resource efficiency, \textit{tenant isolation}, and \textit{tenant provisioning}. The results demonstrate that the proposed architecture successfully manages multiple tenants within a single database and schema without data leakage between tenants, consistently enforces tenant isolation at both the application and API layers, and enables automated \textit{tenant provisioning} without requiring manual configuration. Furthermore, the \textit{shared database} and \textit{shared schema} approach achieves more efficient infrastructure resource utilization than both the non-\textit{multitenant} and \textit{separated database} approaches. Therefore, the proposed architecture can serve as an effective solution for developing \textit{SaaS} applications that support multiple organizations while maintaining infrastructure efficiency, data isolation, and simplified tenant management.

\vspace{1cm}

\noindent\textbf{Keywords:} \textit{Software as a Service} (\textit{SaaS}), \textit{multitenancy}, \textit{shared database}, \textit{shared schema}, \textit{tenant isolation}, \textit{tenant provisioning}.
